---
title: "eda"
author: "Kelsey Schroeder"
date: "August 12, 2015"
output: html_document
---
For more details on using R Markdown see <http://rmarkdown.rstudio.com>.


```{r}
```

You can also embed plots, for example:

```{r, echo=FALSE, message = FALSE}
source('db_connection.r')
library(ggplot2)
library(tidyr)
library(dplyr)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

First, I'll pull the data set from my database.
```{r}
taxi_raw <- query_results("
  SELECT *
 FROM trips
 INNER JOIN fares ON trips.id = fares.trip_id
 ORDER BY trips.hack_license, trips.medallion, trips.pickup_datetime
 LIMIT 1000000;

")

taxi_raw <- taxi_raw[,unique(colnames(taxi_raw))]

```

In order to better analyze the data, let's divide the longitude/latitude pairings into boroughs (this is just an estimation, boundaries aren't perfect):

```{r}
library(dplyr)
taxi_raw <- taxi_raw %>% 
  mutate(pickup_borough = 
    ifelse(pickup_longitude> -74.01 & pickup_longitude< -73.927 & 
             pickup_latitude > 40.701 & pickup_latitude < 40.875, "manhattan", 
    ifelse(pickup_latitude<40.7 & pickup_longitude > -73.85 | 
             pickup_longitude > -73.96 & pickup_longitude < -73.92 & 
                pickup_latitude < 40.704 & pickup_latitude >40.739, "brooklyn", 
    ifelse(pickup_longitude> - 73.927 | pickup_latitude > 40.875, "bronx",
    ifelse(pickup_longitude < -74.04, "staten_island", "queens")))))

taxi_raw <- taxi_raw %>% 
  mutate(dropoff_borough = 
    ifelse(dropoff_longitude> -74.01 & dropoff_longitude< -73.927 & 
             dropoff_latitude > 40.701 & dropoff_latitude < 40.875, "manhattan", 
    ifelse(dropoff_latitude<40.7 & dropoff_longitude > -73.85 | 
             dropoff_longitude > -73.96 & dropoff_longitude < -73.92 & 
                dropoff_latitude < 40.704 & dropoff_latitude >40.739, "brooklyn", 
    ifelse(dropoff_longitude> - 73.927 | dropoff_latitude > 40.875, "bronx",
    ifelse(dropoff_longitude < -74.04, "staten_island", "queens")))))

taxi <- taxi_raw %>% 
  tidyr::separate(pickup_datetime, c("pickup_date", "pickup_time"), 10)
```

How many pickups per borough?

```{r}
borough_total_pickups <- taxi %>% 
  filter(!is.na(pickup_borough) & !is.na(dropoff_borough)) %>% 
  group_by(pickup_borough) %>% 
  summarize(count = n())
```

Let's perform some EDA.  First, let's look into tipping.  What's the distribution of tip percentages? 

TODO: Tried doing a ratio here but it didn't work.  Try again later?

```{r, echo = FALSE}
library(ggplot2)
taxi$tip_amount = as.double(taxi$tip_amount)
taxi$fare_amount = as.double(taxi$fare_amount)


tip_percentages <- taxi %>% 
  dplyr::mutate(tip_percentage = tip_amount/fare_amount*100) %>%
  left_join(borough_total_pickups, by = "pickup_borough") %>% 
  group_by(pickup_borough)
  
ggplot(tip_percentages, aes(tip_percentage, fill=pickup_borough)) + geom_bar(position="dodge", binwidth = .1) + xlim(0,.5)

```

How many trips per day?

```{r, echo = FALSE}
days_of_week <- read.csv("/Users/angelaschroeder/Downloads/days_of_week - Sheet1 (1).csv")

taxi <- taxi %>% 
  left_join(days_of_week, by = "pickup_date")

taxi %>% 
  group_by(weekday) %>% 
  summarize(total_rides = n())

week_levels <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

ggplot(aes(x = factor(weekday, week_levels)), data = taxi) + geom_histogram() + xlab("Day of Week") +
  ylab("Number of Trips") + ggtitle("Trips Per Day")
```

Total revenue each day of the week?

```{r, echo = FALSE}
taxi$total_amount <- as.double(taxi$total_amount) 
taxi %>% 
  group_by(weekday) %>% 
  summarize(revenue = sum(total_amount)) %>% 
  arrange(desc(revenue))

ggplot(taxi, aes(x = factor(weekday, week_levels)), y = revenue) + geom_bar() + 
  xlab("Day of Week") + ylab("Total Revenue ($)") + ggtitle("Revenue Per Day of Week")
```



