---
title: "eda"
author: "Kelsey Schroeder"
date: "August 12, 2015"
output: html_document
---
```{r, echo=FALSE, message = FALSE}
source('db_connection.r')
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
library('rpart') # NOTE: you need to manually install rpart.plot as a separate package
library(rpart.plot)
library(corrgram)
library(prob)
```

First, I'll pull the data set from my database.
```{r}
taxi_raw <- query_results("
  SELECT *
 FROM trips
 INNER JOIN fares ON trips.id = fares.trip_id
 -- ORDER BY trips.hack_license, trips.medallion, trips.pickup_datetime
 LIMIT 1000000;

")

taxi_raw <- taxi_raw[,unique(colnames(taxi_raw))]
```

In order to better analyze the data, let's divide the longitude/latitude pairings into boroughs (this is just an estimation, boundaries aren't perfect):

```{r}
taxi_raw <- taxi_raw %>% 
  mutate(pickup_borough = 
    ifelse(pickup_longitude> -74.01 & pickup_longitude< -73.927 & 
             pickup_latitude > 40.701 & pickup_latitude < 40.875, "manhattan", 
    ifelse(pickup_latitude<40.7 & pickup_longitude > -73.85 | 
             pickup_longitude > -73.96 & pickup_longitude < -73.92 & 
                pickup_latitude < 40.704 & pickup_latitude >40.739, "brooklyn", 
    ifelse(pickup_longitude> - 73.927 | pickup_latitude > 40.875, "bronx",
    ifelse(pickup_longitude < -74.04, "staten_island", "queens")))))

taxi_raw <- taxi_raw %>% 
  mutate(dropoff_borough = 
    ifelse(dropoff_longitude> -74.01 & dropoff_longitude< -73.927 & 
             dropoff_latitude > 40.701 & dropoff_latitude < 40.875, "manhattan", 
    ifelse(dropoff_latitude<40.7 & dropoff_longitude > -73.85 | 
             dropoff_longitude > -73.96 & dropoff_longitude < -73.92 & 
                dropoff_latitude < 40.704 & dropoff_latitude >40.739, "brooklyn", 
    ifelse(dropoff_longitude> - 73.927 | dropoff_latitude > 40.875, "bronx",
    ifelse(dropoff_longitude < -74.04, "staten_island", "queens")))))

taxi_raw <- taxi_raw %>% 
  mutate(
    medallion = as.factor(medallion),
    hack_license = as.factor(hack_license),
    vendor_id = as.factor(vendor_id),
    rate_code = as.factor(rate_code),
    pickup_datetime = ymd_hms(pickup_datetime),
    dropoff_datetime = ymd_hms(dropoff_datetime), # UTC incorrect
    payment_type = as.factor(payment_type),
    fare_amount = as.double(fare_amount),
    surcharge = as.double(surcharge),
    mta_tax = as.double(mta_tax),
    tip_amount = as.double(tip_amount),
    tolls_amount = as.double(tolls_amount),
    total_amount = as.double(total_amount),
    pickup_borough = as.factor(pickup_borough),
    dropoff_borough = as.factor(dropoff_borough),
    trip_time_in_secs = as.integer(trip_time_in_secs),
    trip_distance = as.double(trip_distance),
    passenger_count = as.integer(passenger_count)
  )
```

How many pickups per borough?

```{r}
borough_total_pickups <- taxi_raw %>% 
  dplyr::filter(!is.na(pickup_borough) & !is.na(dropoff_borough)) %>% 
  group_by(pickup_borough) %>% 
  summarize(borough_pickups = n())
```

Let's perform some EDA. How many trips per day of week?

```{r, echo = FALSE, message = FALSE}
days_of_week <- read.csv("days_of_week - Sheet1 (1).csv")

taxi <- taxi_raw %>% 
  dplyr::mutate(tip_percentage = tip_amount/fare_amount*100) %>%
  left_join(borough_total_pickups, by = "pickup_borough") %>% 
  tidyr::separate(pickup_datetime, c("pickup_date", "pickup_time"), " ") %>%
  tidyr::separate(pickup_time, c("pickup_hour", "pickup_min", "pickup_sec"), ":") %>% 
  left_join(days_of_week, by = "pickup_date") %>% 
  mutate(pickup_date = ymd(pickup_date)) %>% 
  mutate(pickup_hour = as.integer(pickup_hour)) %>% 
  mutate(pickup_min = as.integer(pickup_min)) %>% 
  mutate(pickup_sec = as.integer(pickup_sec))

taxi %>% 
  group_by(weekday) %>% 
  summarize(total_rides = n())

week_levels <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

ggplot(aes(x = factor(weekday, week_levels)), data = taxi) + geom_histogram() + xlab("Day of Week") +
  ylab("Number of Trips") + ggtitle("Trips Per Day")
```

How many trips per date?  We can see here the lull on January 21st (MLK Day).  There's
also a dip on Janurary 6th (Sunday), which is the Epiphany, but I'm not convinced that's why we see less taxi traffic.  Also puzzling is the peak on January 26th (Saturday).

```{r, echo = F, message = FALSE}
jan_taxi <- taxi %>% 
  group_by(pickup_date) %>% 
  summarize(trips = n())

ggplot(jan_taxi, aes(x = pickup_date, y = trips, group = 1)) + geom_line(stat= 'identity') +
  ylab("Number of Trips") + xlab("Pickup Date") + ggtitle("Number of Daily Trips in January 2013")
```

Total revenue each day of the week?
TODO: See if this is statistically significant

```{r, echo = FALSE, message = FALSE}
taxi_days <- taxi %>% 
  group_by(pickup_date, weekday) %>% 
  summarize(revenue = sum(total_amount)) %>% 
  arrange(desc(revenue))

week_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

ggplot(taxi_days, aes(x = factor(weekday,week_levels), y = revenue)) + geom_boxplot() + 
  xlab("Day of Week") + ylab("Total Revenue ($)") + ggtitle("Revenue Per Day of Week")
```

It appears that there are, in fact, differences in revenue throughout the week.  Keep in mind that our sample contains just four of each day (totaling 1 month).  I'm particularly interested in Monday and Friday, which seem to be the lowest and highest revenue days, respecively. In order to estimate significance, we can perform a t-test between Monday and Friday and between each of these days and Saturday, which looks to be about the middle.


```{r}
# new data frame for each day of the week we're analyzing
taxi_days_M <- taxi_days %>% 
  dplyr::filter(weekday == "Monday")
taxi_days_F <- taxi_days %>% 
  dplyr::filter(weekday == "Friday")
taxi_days_S <- taxi_days %>% 
  dplyr::filter(weekday == "Saturday")

t.test(taxi_days_M$revenue, taxi_days_F$revenue)
t.test(taxi_days_M$revenue, taxi_days_S$revenue)
t.test(taxi_days_F$revenue, taxi_days_S$revenue)
```

Because the p-value of the t-test for Friday/Saturday is >,05, we can conclude that the revenue on those two days is not statistically significant.  However, both the Monday/Friday and Monday/Saturday tests are significant, indicating that the Monday revenue is not the same as the Friday and Saturday revenues.  

Now I'll explore the toll situation in the city.  High tolls for Staten Island!

```{r}
taxi %>% 
  dplyr::filter(tolls_amount>0) %>% 
  group_by(pickup_borough, dropoff_borough) %>% 
  summarize(num_rides_w_tolls = n(), average_toll = mean(tolls_amount)) %>% 
  ungroup() %>% 
  ungroup() %>% 
  arrange(desc(average_toll)) # can also order by number of toll rides
```

How about trip length?

```{r, echo = FALSE, message = FALSE}
taxi_length <- taxi %>% 
  select(trip_distance, trip_time_in_secs) %>% 
  dplyr::filter(trip_distance>0) 

taxi_length$trip_distance <- round(taxi_length$trip_distance)
taxi_length$trip_time_in_secs <- round(taxi_length$trip_time_in_secs, digits = -1)

ggplot(taxi_length, aes(x=trip_distance, y= trip_time_in_secs)) +
    geom_point(shape=19, alpha=1/50) + xlim(0,30)

summary(taxi_length) 
```

Looks like the vast majority of trips are within 25 miles and 50 minutes.  Mean distance is 3.39 miles, mean time is about 13 minutes (802 seconds).

Can we find any trends with trips to airports? Perhaps there's a flat rate fare.  Here's a histogram of fare amounts (not counting surcharges, tip, tolls, etc.).

```{r, echo = FALSE}
ggplot(taxi, aes(x = fare_amount)) + geom_histogram(binwidth = 1) + xlim(0,57)
```

Notice the peak at $52.  Perhaps this is a flat rate?

```{r}
taxi52 <- taxi %>% 
  dplyr::filter(fare_amount == 52)

ggplot(taxi52, aes(x = dropoff_longitude, y = dropoff_latitude)) + geom_point(alpha = 1/10) + 
  xlim(-74.2,-73.6) + ylim(40.5,41)
```

Here we see two clusters.  Let's examine more closely.

```{r}
# Checking out the cluster to the right
ggplot(taxi52, aes(x = pickup_longitude, y = pickup_latitude)) + geom_point(alpha = 1/10) + 
  xlim(-74.05,-73.75)+ ylim(40.5,41)
```

And would you look at that!  The cluster to the right is at approximately -73.78 longitude and 40.645 latitude, landing us right on top of JFK airport.  If we look at the pickup data for the other point in the dropoff graph, we find that these customers were picked up at the airport and brough to various places in NYC.  The internet confirms that this ride is a flat rate either way.

What if we look at general pickup/dropoff locations?  Will we find any especially popular spots other than the airport?

```{r}
ggplot(taxi, aes( x = pickup_longitude, y = pickup_latitude)) + geom_point(alpha = 1/500) + 
  xlim(-74.01,-73.85) + ylim(40.7,40.85) + theme_bw() # having trouble setting this alpha to the correct value. want to be able to pick out specific long/lat points. Cool map of Manhattan for now!

ggplot(taxi, aes( x = dropoff_longitude, y = dropoff_latitude)) + geom_point(alpha = 1/100) + xlim(-74.4,-73.4) + ylim(40.5,41) + theme_bw()
```

By zooming in on this graph and identifying landmarks based on longitude/latitude data, I was able to identify the following as popular pick up and drop off locations:
* LaGuardia Airport
* Times Square
* Penn Station/MSG
* Columbus Circle
* Car Inspection Station in Brooklyn
* Dial 7 car/limo service in Queens
* Brooklyn Driving School
* Various auto shops

Let's look at tipping.  This data set only contains tip data for rides paid with cards, so we'll limit our data set to those trips only.

```{r}
tip_analysis <- taxi %>% 
  filter(payment_type=="CRD") %>% 
  filter(tip_percentage<=100)

ggplot(data = tip_analysis, aes(x = fare_amount, y = tip_percentage)) + 
  geom_point(alpha = 1/10) + xlim(0,100)

ggplot(data = tip_analysis, aes(x = tip_percentage)) + geom_histogram() + xlim(0,50)
```

The first graph yields 2 main conclusions: 
1. Customers are using the buttons on the card payment screen for 20, 25, and 30% tips (these are the horizontal lines in the graph).
2. Customers pay a smaller percentage when the rate is high (i.e. If I'm already spending $50 on a taxi, I don't want to spend another $15 on the tip).  

We can also see the vertical line at the $52 fare (many different tips for airport trips).

Let's run a correlogram to see if there are any surprising relationships between our variables.

```{r}
taxi_num <- taxi %>% 
  dplyr::mutate(tip_percentage = tip_amount/fare_amount*100) %>% 
  select(vendor_id, rate_code, passenger_count, trip_time_in_secs, trip_distance, 
         pickup_longitude, pickup_latitude, dropoff_longitude, dropoff_latitude,
         fare_amount, surcharge, mta_tax, tip_amount, tolls_amount, total_amount,
         tip_percentage)

corrgram(taxi_num, order = TRUE, upper.panel = panel.pie)
```

Nothing too interesting here... the fare, trip time, trip distance, tip amount total amount, etc. are all positively correlated.  The pickup longitude and latitude seem to only affect each other, which makes sense since in NYC longitudes are negative while latitudes are positive.  Correlograms have been helpful in analyses of other data sets, but here this method just illuminates the lack of surprising correlations in our data.

Let's take a look at trip efficiency.  Here I'll define efficiency as trip distance/trip time (this will give us the average mph of the trip). 

```{r}
taxi_efficiency <- taxi %>% 
  dplyr::filter(trip_distance>0, trip_time_in_secs>0) %>% 
  mutate(mph = trip_distance/(trip_time_in_secs/3600)) %>% 
  arrange(mph)

ggplot(data = taxi_efficiency, aes(x = pickup_hour, y = mph)) + 
  geom_point(alpha = 1/510) + ylim(0,30) + geom_smooth(se = F)
```

Looks like the efficiency peaks at around 4:30am but remains pretty consistent from 7am-7pm.  We see higher efficiency outside of work hours, as expected.

```{r}
taxi_inefficient <- taxi_efficiency %>% 
  dplyr::filter(mph< 2, mph >.01, pickup_longitude!=0, pickup_latitude!=0)

ggplot(data = taxi_inefficient) + 
  geom_point(aes(y = pickup_latitude, x = pickup_longitude),color = "blue", alpha = 1/25) + 
  xlim(-74.02,-73.94) + ylim(40.7,40.8) + 
  geom_point(aes(x = dropoff_longitude, y = dropoff_latitude), color = "red", alpha = 1/25)

```

Perhaps we can use machine learning to predict the length of a trip.

```{r}
efficient_ml <- taxi_efficiency %>% 
  mutate(trip_time_in_mins = trip_time_in_secs/60) %>% 
  dplyr::filter(dropoff_latitude<41, pickup_latitude<41) %>% 
  select(rate_code, pickup_date, pickup_hour, passenger_count, pickup_longitude, pickup_latitude,
         dropoff_latitude, dropoff_longitude, pickup_borough, dropoff_borough, payment_type,
         surcharge, tolls_amount,weekday, trip_time_in_mins)

fit <- rpart(trip_time_in_mins ~ ., data = efficient_ml, method = "anova")
prp(fit)
```



