---
title: "eda"
author: "Kelsey Schroeder"
date: "August 12, 2015"
output: html_document
---
For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

```{r, echo=FALSE, message = FALSE}
source('db_connection.r')
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
library('rpart') # NOTE: you need to manually install rpart.plot as a separate package
library('rattle')
library(corrgram)
```

First, I'll pull the data set from my database.
```{r}
taxi_raw <- query_results("
  SELECT *
 FROM trips
 INNER JOIN fares ON trips.id = fares.trip_id
 -- ORDER BY trips.hack_license, trips.medallion, trips.pickup_datetime
 LIMIT 1000000;

")

taxi_raw <- taxi_raw[,unique(colnames(taxi_raw))]
```

In order to better analyze the data, let's divide the longitude/latitude pairings into boroughs (this is just an estimation, boundaries aren't perfect):

```{r}
taxi_raw <- taxi_raw %>% 
  mutate(pickup_borough = 
    ifelse(pickup_longitude> -74.01 & pickup_longitude< -73.927 & 
             pickup_latitude > 40.701 & pickup_latitude < 40.875, "manhattan", 
    ifelse(pickup_latitude<40.7 & pickup_longitude > -73.85 | 
             pickup_longitude > -73.96 & pickup_longitude < -73.92 & 
                pickup_latitude < 40.704 & pickup_latitude >40.739, "brooklyn", 
    ifelse(pickup_longitude> - 73.927 | pickup_latitude > 40.875, "bronx",
    ifelse(pickup_longitude < -74.04, "staten_island", "queens")))))

taxi_raw <- taxi_raw %>% 
  mutate(dropoff_borough = 
    ifelse(dropoff_longitude> -74.01 & dropoff_longitude< -73.927 & 
             dropoff_latitude > 40.701 & dropoff_latitude < 40.875, "manhattan", 
    ifelse(dropoff_latitude<40.7 & dropoff_longitude > -73.85 | 
             dropoff_longitude > -73.96 & dropoff_longitude < -73.92 & 
                dropoff_latitude < 40.704 & dropoff_latitude >40.739, "brooklyn", 
    ifelse(dropoff_longitude> - 73.927 | dropoff_latitude > 40.875, "bronx",
    ifelse(dropoff_longitude < -74.04, "staten_island", "queens")))))

taxi_raw <- taxi_raw %>% 
  mutate(
    medallion = as.factor(medallion),
    hack_license = as.factor(hack_license),
    vendor_id = as.factor(vendor_id),
    rate_code = as.factor(rate_code),
    pickup_datetime = ymd_hms(pickup_datetime),
    dropoff_datetime = ymd_hms(dropoff_datetime), # UTC incorrect
    payment_type = as.factor(payment_type),
    fare_amount = as.double(fare_amount),
    surcharge = as.double(surcharge),
    mta_tax = as.double(mta_tax),
    tip_amount = as.double(tip_amount),
    tolls_amount = as.double(tolls_amount),
    total_amount = as.double(total_amount),
    pickup_borough = as.factor(pickup_borough),
    dropoff_borough = as.factor(dropoff_borough),
    trip_time_in_secs = as.integer(trip_time_in_secs),
    trip_distance = as.double(trip_distance),
    passenger_count = as.integer(passenger_count)
  )
```

How many pickups per borough?

```{r}
borough_total_pickups <- taxi_raw %>% 
  filter(!is.na(pickup_borough) & !is.na(dropoff_borough)) %>% 
  group_by(pickup_borough) %>% 
  summarize(borough_pickups = n())
```

Let's perform some EDA.  First, let's look into tipping.  What's the distribution of tip percentages? 

TODO: Tried doing a ratio here but it didn't work.  Try again later?

```{r, echo = FALSE}
days_of_week <- read.csv("days_of_week - Sheet1 (1).csv")

taxi <- taxi_raw %>% 
  dplyr::mutate(tip_percentage = tip_amount/fare_amount*100) %>%
  left_join(borough_total_pickups, by = "pickup_borough") %>% 
  tidyr::separate(pickup_datetime, c("pickup_date", "pickup_time"), " ") %>%
  tidyr::separate(pickup_time, c("pickup_hour", "pickup_min", "pickup_sec"), ":") %>% 
  left_join(days_of_week, by = "pickup_date") %>% 
  mutate(pickup_date = ymd(pickup_date))
  
ggplot(taxi, aes(tip_percentage, fill=pickup_borough)) + geom_bar(position="dodge", binwidth = .1) + xlim(0,.5)
```

How many trips per day of week?

```{r, echo = FALSE}
taxi %>% 
  group_by(weekday) %>% 
  summarize(total_rides = n())

week_levels <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

ggplot(aes(x = factor(weekday, week_levels)), data = taxi) + geom_histogram() + xlab("Day of Week") +
  ylab("Number of Trips") + ggtitle("Trips Per Day")
```

How many trips per date?  We can see here the lull on January 21st (MLK Day).  There's
also a dip on Janurary 6th, which is the Epiphany, but I'm not convinced that's why we see less taxi traffic.  Also puzzling is the peak on January 26th (Saturday).

```{r, echo = F}
jan_taxi <- taxi_days %>% 
  group_by(pickup_date) %>% 
  summarize(trips = n())

ggplot(jan_taxi, aes(x = pickup_date, y = trips, group = 1)) + geom_line(stat= 'identity')
```

Total revenue each day of the week?
TODO: See if this is statistically significant

```{r, echo = FALSE}
taxi %>% 
  group_by(weekday) %>% 
  summarize(revenue = sum(total_amount)) %>% 
  arrange(desc(revenue))

ggplot(taxi, aes(x = factor(weekday, week_levels)), y = revenue) + geom_bar() + 
  xlab("Day of Week") + ylab("Total Revenue ($)") + ggtitle("Revenue Per Day of Week")
```

Checking out the toll situation in the city.  High tolls for Staten Island!

```{r}
taxi %>% 
  filter(tolls_amount>0) %>% 
  group_by(pickup_borough, dropoff_borough) %>% 
  summarize(num_rides_w_tolls = n(), average_toll = mean(tolls_amount)) %>% 
  ungroup() %>% 
  ungroup() %>% 
  arrange(desc(average_toll)) # can also order by number of toll rides
```

How about trip length?

```{r}
taxi_length <- taxi %>% 
  select(trip_distance, trip_time_in_secs) %>% 
  filter(trip_distance>0) 

taxi_length$trip_distance <- round(taxi_length$trip_distance)
taxi_length$trip_time_in_secs <- round(taxi_length$trip_time_in_secs, digits = -1)

ggplot(taxi_length, aes(x=trip_distance, y= trip_time_in_secs)) +
    geom_point(shape=19, alpha=1/10)

summary(taxi_length) 
```

Looks like the vast majority of trips are within 25 miles and 50 minutes.  This is expected.  Mean distance is 3.39 miles, mean time is about 13 minutes (802 seconds).

Can we find any trends with trips to airports? Perhaps there's a flat rate fare.  Here's a histogram of fare amounts (not counting surcharges, tip, tolls, etc.).

```{r, echo = FALSE}
ggplot(taxi, aes(x = fare_amount)) + geom_histogram(binwidth = 1) + xlim(0,57)
```

Notice the peak at $52.  Perhaps this is a flat rate?

```{r}
taxi52 <- taxi %>% 
  filter(fare_amount == 52)

ggplot(taxi52, aes(x = dropoff_longitude, y = dropoff_latitude)) + geom_point(alpha = 1/10) + 
  xlim(-74.2,-73.6) + ylim(40.5,41)
```

Here we see two clusters.  Let's examine more closely.

```{r}
# Checking out the cluster to the right
ggplot(taxi52, aes(x = dropoff_longitude, y = dropoff_latitude)) + geom_point(alpha = 1/10) + 
  xlim(-73.9,-73.7) + ylim(40.635,40.65) # JFK
  
ggplot(taxi52, aes(x = dropoff_longitude, y = dropoff_latitude)) + geom_point(alpha = 1/10) + 
  xlim(-74,-73.97) + ylim(40.745,40.775) # unclear

ggplot(taxi52, aes(x = pickup_longitude, y = pickup_latitude)) + geom_point(alpha = 1/10) + 
  xlim(-74.05,-73.75)+ ylim(40.5,41)

taxi52_pickups <- taxi52 %>% 
  filter(dropoff_longitude < -73.97 & dropoff_longitude > -74 & dropoff_latitude > 40.745 &
           dropoff_latitude < 40.775)

ggplot(taxi52_pickups, aes(x = pickup_longitude, y = pickup_latitude)) + geom_point(alpha = 1/10) +
  xlim(-75,-72.5) + ylim(40.5,41)
```

And would you look at that!  The cluster to the right is at approximately -73.78 longitude and 40.645 latitude, landing us right on top of JFK airport.  If we look at the pickup data for the other point in the dropoff graph, we find that these customers were picked up at the airport and brough to various places in NYC.  The internet confirms that this ride is a flat rate either way.

What if we look at general pickup/dropoff locations?  Will we find any especially popular spots other than the airport?

```{r}
ggplot(taxi, aes( x = pickup_longitude, y = pickup_latitude)) + geom_point(alpha = 1/500) + 
  xlim(-74.01,-73.85) + ylim(40.7,40.85) + theme_bw() # having trouble setting this alpha to the correct value. want to be able to pick out specific long/lat points. Cool map of Manhattan for now!

ggplot(taxi, aes( x = dropoff_longitude, y = dropoff_latitude)) + geom_point(alpha = 1/100) + xlim(-74.4,-73.4) + ylim(40.5,41) + theme_bw()
```

By zooming in on this graph and identifying landmarks based on longitude/latitude data, I was able to identify the following as popular pick up and drop off locations:
* LaGuardia Airport
* Times Square
* Penn Station/MSG
* Columbus Circle
* Car Inspection Station in Brooklyn
* Dial 7 car/limo service in Queens
* Brooklyn Driving School
* Eight Strings and a Whistle
* Various auto shops

Now let's try some machine learning on this data set.

```{r}

tip_percentages <- na.omit(tip_percentages)
tip_percentages <- tip_percentages %>%
  select(rate_code
         , passenger_count
         , trip_time_in_secs
         , trip_distance
         , payment_type
         , fare_amount
         , surcharge
         , mta_tax
         , tolls_amount
         , total_amount
         , pickup_borough
         , tip_percentage
         )

smp_size <- floor(0.75 * nrow(tip_percentages))
set.seed(123)
train_ind <- sample(seq_len(nrow(tip_percentages)), size = smp_size)

train <- tip_percentages[train_ind, ]
test <- tip_percentages[-train_ind, ]

rfit <- rpart(tip_percentage ~ . , data=train) 
rpart.plot(rfit)
text(rfit)

rattle::fancyRpartPlot(rfit) # not super correlated, cash tips aren't recorded :(

#tree.pred = predict(tree.reg, test, type = "class")
#with(test, table(tree.pred,High))

library(class)
knn.pred = knn(train_num,test_num, train_num1$High, k = 3)
table(knn.pred, test_num1$High) # 87.5% accuracy
mean(knn.pred == test_num1$High)
plot(knn.pred)
```

Maybe a correlogram can shed some more light on the relationships between our variables.

```{r}
taxi_num <- taxi %>% 
  dplyr::mutate(tip_percentage = tip_amount/fare_amount*100) %>% 
  select(vendor_id, rate_code, passenger_count, trip_time_in_secs, trip_distance, 
         pickup_longitude, pickup_latitude, dropoff_longitude, dropoff_latitude,
         fare_amount, surcharge, mta_tax, tip_amount, tolls_amount, total_amount,
         tip_percentage)

corrgram(taxi_num, order = TRUE, upper.panel = panel.pie)
```

Nothing too interesting here... the fare, trip time, trip distance, tip amount total amount, etc. are all positively correlated.  The pickup longitude and latitude seem to only affect each other, which makes sense since in NYC longitudes are negative while latitudes are positive.  Correlograms have been helpful in analyses of other data sets, but here this method just illuminates the lack of strong correlations in our data.

I'm interested in the revenues/profit strategies of specific cabbies.  Let's take a look at this data, grouped by individual taxi.

```{r}
taxi_indiv <- taxi %>% 
  group_by(hack_license, vendor_id, rate_code, pickup_date, pickup_hour, payment_type, dropoff_datetime,
           pickup_borough, dropoff_borough) %>% 
  summarize(total_distance = sum(trip_distance)
            , total_time = sum(trip_time_in_secs)
            , total_num_passengers = sum(passenger_count)
            , total_tips = sum(tip_amount)
            , total_revenue = sum(total_amount)
            , num_different_cars = n_distinct(medallion)
  ) %>% 
  ungroup() 
```

Now we have a data set for individual cabs.  Yay!  Let's look at a few different things, starting with
the total revenue per cab.
TODO: Use whole data set for this.  The info isn't really accurate without it.  If we can't handle this, could potentially limit date to 2 weeks, 1 week, x total hack licenses, etc.

```{r}
taxi_indiv_rev <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_revenue = sum(total_revenue)) %>% 
  arrange(desc(total_revenue))

ggplot(taxi_indiv_rev, aes(x = total_revenue)) + geom_histogram(binwidth= 20)
```

How about total number of rides, or passengers?

```{r}
taxi_indiv_pass <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(num_passengers = sum(total_num_passengers), num_trips = n_distinct(dropoff_datetime))

ggplot(taxi_indiv_pass) + geom_histogram(aes(num_passengers)) + xlim(0,400) # will need to change these limits for full data set
ggplot(taxi_indiv_pass) + geom_histogram(aes(num_trips)) + xlim(0,100)
```

Total number of different cars driven by cabbie: THIS IS WRONG
TODO: Figure out why some cabbies apparently drive 132 different cars in a month.
```{r}
taxi_indiv_cars <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_cars = sum(num_different_cars)) %>% 
  arrange(desc(total_cars))
```
