---
title: "Untitled"
author: "Kelsey Schroeder"
date: "August 19, 2015"
output: html_document
---

```{r}
source('db_connection.r')
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
library(corrgram)
library(NMF)
taxi_strat_raw <- query_results("
  SELECT *
 FROM trips
 INNER JOIN fares ON trips.id = fares.trip_id
 WHERE trips.hack_license IN (
   SELECT hack_license FROM trips ORDER BY random() LIMIT 967
  )
;")

taxi_strat_raw <- taxi_strat_raw[,unique(colnames(taxi_strat_raw))]

taxi_strat_raw <- taxi_strat_raw %>% 
  mutate(pickup_borough = 
    ifelse(pickup_longitude> -74.01 & pickup_longitude< -73.927 & 
             pickup_latitude > 40.701 & pickup_latitude < 40.875, "manhattan", 
    ifelse(pickup_latitude<40.7 & pickup_longitude > -73.85 | 
             pickup_longitude > -73.96 & pickup_longitude < -73.92 & 
                pickup_latitude < 40.704 & pickup_latitude >40.739, "brooklyn", 
    ifelse(pickup_longitude> - 73.927 | pickup_latitude > 40.875, "bronx",
    ifelse(pickup_longitude < -74.04, "staten_island", "queens")))))

taxi_strat_raw <- taxi_strat_raw %>% 
  mutate(dropoff_borough = 
    ifelse(dropoff_longitude> -74.01 & dropoff_longitude< -73.927 & 
             dropoff_latitude > 40.701 & dropoff_latitude < 40.875, "manhattan", 
    ifelse(dropoff_latitude<40.7 & dropoff_longitude > -73.85 | 
             dropoff_longitude > -73.96 & dropoff_longitude < -73.92 & 
                dropoff_latitude < 40.704 & dropoff_latitude >40.739, "brooklyn", 
    ifelse(dropoff_longitude> - 73.927 | dropoff_latitude > 40.875, "bronx",
    ifelse(dropoff_longitude < -74.04, "staten_island", "queens")))))

taxi_strat_raw <- taxi_strat_raw %>% 
  mutate(
    medallion = as.factor(medallion),
    hack_license = as.factor(hack_license),
    vendor_id = as.factor(vendor_id),
    rate_code = as.factor(rate_code),
    pickup_datetime = ymd_hms(pickup_datetime),
    dropoff_datetime = ymd_hms(dropoff_datetime), # UTC incorrect
    payment_type = as.factor(payment_type),
    fare_amount = as.double(fare_amount),
    surcharge = as.double(surcharge),
    mta_tax = as.double(mta_tax),
    tip_amount = as.double(tip_amount),
    tolls_amount = as.double(tolls_amount),
    total_amount = as.double(total_amount),
    pickup_borough = as.factor(pickup_borough),
    dropoff_borough = as.factor(dropoff_borough),
    trip_time_in_secs = as.integer(trip_time_in_secs),
    trip_distance = as.double(trip_distance),
    passenger_count = as.integer(passenger_count)
  )
borough_total_pickups <- taxi_strat_raw %>% 
  dplyr::filter(!is.na(pickup_borough) & !is.na(dropoff_borough)) %>% 
  group_by(pickup_borough) %>% 
  summarize(borough_pickups = n())
days_of_week <- read.csv("days_of_week - Sheet1 (1).csv")

taxi_strat <- taxi_strat_raw %>% 
  dplyr::mutate(tip_percentage = tip_amount/fare_amount*100) %>%
  left_join(borough_total_pickups, by = "pickup_borough") %>% 
  tidyr::separate(pickup_datetime, c("pickup_date", "pickup_time"), " ") %>%
  tidyr::separate(pickup_time, c("pickup_hour", "pickup_min", "pickup_sec"), ":") %>% 
  left_join(days_of_week, by = "pickup_date") %>% 
  mutate(pickup_date = ymd(pickup_date)) %>% 
  mutate(pickup_hour = as.integer(pickup_hour)) %>% 
  mutate(pickup_min = as.integer(pickup_min)) %>% 
  mutate(pickup_sec = as.integer(pickup_sec))
  
taxi_indiv <- taxi_strat %>% 
  group_by(hack_license, vendor_id, rate_code, pickup_date, pickup_hour, payment_type, 
           pickup_borough, dropoff_borough) %>% 
  summarize(total_distance = sum(trip_distance)
            , total_time = sum(trip_time_in_secs)
            , total_num_passengers = sum(passenger_count)
            , total_tips = sum(tip_amount)
            , total_revenue = sum(total_amount)
            , num_different_cars = n_distinct(medallion)
            , total_trips = n_distinct(dropoff_datetime)
  ) %>% 
  ungroup() 
```

I'm interested in the revenues/profit strategies of specific cabbies.  Let's take a look at this data, grouped by individual taxi.

Let's look at a few different things, starting with
the total revenue per cab.

```{r}
taxi_indiv_rev <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_revenue = sum(total_revenue)) %>% 
  arrange(desc(total_revenue))

ggplot(taxi_indiv_rev, aes(x = total_revenue)) + geom_histogram(binwidth= 500)
```

How about total number of passengers/trips?

```{r}
taxi_indiv_pass <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(num_passengers = sum(total_num_passengers), num_trips = sum(total_trips))

ggplot(taxi_indiv_pass) + geom_histogram(aes(num_passengers)) + xlim(0,4000) # will need to change these limits for full data set
ggplot(taxi_indiv_pass) + geom_histogram(aes(num_trips)) + xlim(0,1000)
```

Total number of different cars driven by cabbie: THIS IS WRONG
TODO: Figure out why some cabbies apparently drive 132 different cars in a month.
```{r}
taxi_indiv_cars <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_cars = sum(num_different_cars)) %>% 
  arrange(desc(total_cars))
```

Cabbies who are tipped the best (relative to their revenue):

```{r}
taxi_indiv_tips <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_revenue = sum(total_revenue), total_tips = sum(total_tips)) %>% 
  mutate(tip_rate = total_tips/total_revenue)

ggplot(taxi_indiv_tips, aes(tip_rate)) + geom_histogram(binwidth = .005) 
```

Number of days worked in the month:
```{r}
taxi_indiv_days <- taxi_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_days_worked = n_distinct(pickup_date)) %>% 
  arrange(desc(total_days_worked))

ggplot(taxi_indiv_days, aes(total_days_worked)) + geom_histogram()
```

Wow. Looks like many of our cabbies worked every single day of January.  That's dedication!

Maybe we can figure out what shift the cabbies work by labeling their pickup times with a broader category tag.  This assumes a shift change at 5pm (assumptions based on IQuantNY analysis).

```{r}
taxi_indiv_shifts <- taxi_indiv %>% 
  mutate(tod = ifelse(pickup_hour>5 & pickup_hour<17, "morning_afternoon", "evening_night")) %>% 
  group_by(hack_license, tod) %>% 
  summarize(count = n()) %>% 
  spread(tod, count, fill = 0) %>% 
  mutate(morning_per_evening = morning_afternoon/evening_night)

ggplot(taxi_indiv_shifts) + geom_histogram(aes(evening_night), binwidth = 1)
ggplot(taxi_indiv_shifts) + geom_histogram(aes(morning_afternoon), binwidth = 1)
ggplot(taxi_indiv_shifts) + geom_histogram(aes(morning_per_evening), binwidth = .1) +
  xlim(0,5)
```

Are there certain drivers that have only cash/only card payments? Maybe drive cars without card capabilities?

Most cabbies are a little less than 50% cash.  There are 3 who have mroe than 10 trips and all are cash.  None have 0% cash.
```{r}
taxi_indiv_payment <- taxi_indiv %>% 
  group_by(hack_license, payment_type) %>% 
  summarize(count = n()) %>% 
  spread(payment_type, count, fill = 0) %>% 
  mutate(total = CRD + CSH + DIS + NOC + UNK) %>% 
  mutate(CRD = CRD/total, CSH = CSH/total, DIS = DIS/total, NOC = NOC/total, UNK = UNK/total) %>% 
  arrange(desc(CSH)) %>% 
  filter(total>=10)

ggplot(taxi_indiv_payment) + geom_histogram(aes(CSH), binwidth = .05)
```

Let's look at cabbies' driving schedules.  We can use NMF to find relationships between different pickup hours of trips.
```{r}
taxi_indiv_nmf <- taxi_indiv %>% 
  group_by(hack_license, pickup_hour) %>% 
  summarize(count = n()) %>% 
  mutate(count = ifelse(count>5,1,0)) %>% 
  dplyr::filter(!is.na(count), !is.na(hack_license), !is.na(pickup_hour)) %>% 
  spread(pickup_hour, count, fill = 0) %>% 
  select(-hack_license)

taxi_indiv_nmf <- as.matrix(taxi_indiv_nmf)
taxi_indiv_nmf <- na.omit(taxi_indiv_nmf)

nmf_taxi <- nmf(as.matrix(taxi_indiv_nmf), rank = 4, nrun = 10)
```


