---
title: "profit_strategies"
author: "Kelsey Schroeder"
date: "8/19/2015"
output: html_document
---

Here, let's investigate profit strategies for individual taxi drivers.   

```{r setup, message = FALSE, results = 'hide'}
source('db_connection.r')
library(markdown)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
library(corrgram)
library(rpart.plot)
library(gridExtra)
```

```{r}
taxi_strat_raw <- query_results("
 SELECT setseed(.5); 
 SELECT *
 FROM trips
 INNER JOIN fares ON trips.id = fares.trip_id
 WHERE trips.hack_license IN (
   SELECT hack_license FROM trips ORDER BY random() LIMIT 967
  )
;")

taxi_strat_raw <- taxi_strat_raw[,unique(colnames(taxi_strat_raw))]

taxi_strat_raw <- taxi_strat_raw %>% 
  mutate(pickup_borough = 
    ifelse(pickup_longitude> -74.01 & pickup_longitude< -73.927 & 
             pickup_latitude > 40.701 & pickup_latitude < 40.875, "manhattan", 
    ifelse(pickup_latitude<40.7 & pickup_longitude > -73.85 | 
             pickup_longitude > -73.96 & pickup_longitude < -73.92 & 
                pickup_latitude < 40.704 & pickup_latitude >40.739, "brooklyn", 
    ifelse(pickup_longitude> - 73.927 | pickup_latitude > 40.875, "bronx",
    ifelse(pickup_longitude < -74.04, "staten_island", "queens")))))

taxi_strat_raw <- taxi_strat_raw %>% 
  mutate(dropoff_borough = 
    ifelse(dropoff_longitude> -74.01 & dropoff_longitude< -73.927 & 
             dropoff_latitude > 40.701 & dropoff_latitude < 40.875, "manhattan", 
    ifelse(dropoff_latitude<40.7 & dropoff_longitude > -73.85 | 
             dropoff_longitude > -73.96 & dropoff_longitude < -73.92 & 
                dropoff_latitude < 40.704 & dropoff_latitude >40.739, "brooklyn", 
    ifelse(dropoff_longitude> - 73.927 | dropoff_latitude > 40.875, "bronx",
    ifelse(dropoff_longitude < -74.04, "staten_island", "queens")))))

taxi_strat_raw <- taxi_strat_raw %>% 
  mutate(
    medallion = as.factor(medallion),
    hack_license = as.factor(hack_license),
    vendor_id = as.factor(vendor_id),
    rate_code = as.factor(rate_code),
    pickup_datetime = ymd_hms(pickup_datetime),
    dropoff_datetime = ymd_hms(dropoff_datetime), # UTC incorrect
    payment_type = as.factor(payment_type),
    fare_amount = as.double(fare_amount),
    surcharge = as.double(surcharge),
    mta_tax = as.double(mta_tax),
    tip_amount = as.double(tip_amount),
    tolls_amount = as.double(tolls_amount),
    total_amount = as.double(total_amount),
    pickup_borough = as.factor(pickup_borough),
    dropoff_borough = as.factor(dropoff_borough),
    trip_time_in_secs = as.integer(trip_time_in_secs),
    trip_distance = as.double(trip_distance),
    passenger_count = as.integer(passenger_count)
  )
borough_total_pickups <- taxi_strat_raw %>% 
  filter(!is.na(pickup_borough) & !is.na(dropoff_borough)) %>% 
  group_by(pickup_borough) %>% 
  summarize(borough_pickups = n())
days_of_week <- read.csv("days_of_week - Sheet1 (1).csv")

taxi_strat <- taxi_strat_raw %>% 
  dplyr::mutate(tip_percentage = tip_amount/fare_amount*100) %>%
  left_join(borough_total_pickups, by = "pickup_borough") %>% 
  mutate(pickup_change = pickup_datetime) %>% 
  tidyr::separate(pickup_change, c("pickup_date", "pickup_time"), " ") %>%
  tidyr::separate(pickup_time, c("pickup_hour", "pickup_min", "pickup_sec"), ":") %>% 
  left_join(days_of_week, by = "pickup_date") %>% 
  mutate(pickup_date = ymd(pickup_date)) %>% 
  mutate(pickup_hour = as.integer(pickup_hour)) %>% 
  mutate(pickup_min = as.integer(pickup_min)) %>% 
  mutate(pickup_sec = as.integer(pickup_sec))
  
strat_indiv <- taxi_strat %>% 
  group_by(hack_license, vendor_id, rate_code, pickup_date, pickup_hour, payment_type, 
           pickup_borough, dropoff_borough) %>% 
  summarize(total_distance = sum(trip_distance)
            , total_time = sum(trip_time_in_secs)
            , total_num_passengers = sum(passenger_count)
            , total_tips = sum(tip_amount)
            , total_revenue = sum(total_amount)
            , num_different_cars = n_distinct(medallion)
            , total_trips = n_distinct(dropoff_datetime)
  ) %>% 
  ungroup() 
```

Does more hours always mean more revenue?

```{r}
strat_indiv_rev_time <- strat_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_time= sum(total_time), total_revenue = sum(total_revenue))

ggplot(strat_indiv_rev_time, aes(x = total_time, y = total_revenue)) + geom_point()
```

Looks like the relationship is pretty linear (as expected).  Can we find some other correlations in our data frame?

```{r}
corrgram(strat_indiv, order = TRUE, upper.panel = panel.pie)
```

Trips to revenue?
```{r}
strat_indiv_trips <- strat_indiv %>% 
  group_by(hack_license) %>% 
  summarize(total_revenue = sum(total_revenue), total_trips = sum(total_trips))

ggplot(strat_indiv_trips, aes(x = total_trips, y = total_revenue)) + geom_point()
fit <- lm(total_revenue ~ total_trips, data = strat_indiv_trips)
summary(fit)
```

This is again a pretty linear relationship.  I'm interested in the outliers, in particular drivers who had fewer trips but more revenue, or fewer hours but more revenue.

Let's compute the total number of hours each cabbie spends in service (not necessarily on the meter).

```{r}
trip_times <- taxi_strat %>% 
  group_by(hack_license, pickup_date) %>% 
  summarize(secs_on_meter = sum(trip_time_in_secs), total_amount = sum(total_amount))

taxi_empty <- taxi_strat %>% 
  select(hack_license, pickup_date, pickup_datetime, dropoff_datetime) %>% 
  arrange(hack_license, pickup_date, pickup_datetime) 
# not using mutate here because my version of dplyr has a bug with mutate and lubridate
taxi_empty$empty_time <- as.integer(taxi_empty$pickup_datetime - lag(taxi_empty$dropoff_datetime))

taxi_empty <- taxi_empty %>% 
  mutate(empty_time = ifelse(is.na(empty_time)|empty_time<0|empty_time>3600,0,empty_time)) %>%  
  group_by(hack_license, pickup_date) %>% 
  summarize(total_empty_time = sum(empty_time)) %>% 
  left_join(trip_times, by = c("hack_license" = "hack_license", 
                               "pickup_date" = "pickup_date")) %>% 
  mutate(total_hours = (total_empty_time + secs_on_meter)/3600)

ggplot(data = taxi_empty, aes(x = total_hours)) + geom_histogram() + xlim(0,20)
summary(taxi_empty$total_hours)
```

So, on any given day, we can see the number of hours worked by a cabbie. The mean workday consists of 7.22 hours (this does not include any meter breaks longer than an hour).

Let's look at efficiency from this point of view.
```{r}
taxi_empty_month <- taxi_empty %>% 
  group_by(hack_license) %>% 
  summarize(total_empty_time = sum(total_empty_time), secs_on_meter = sum(secs_on_meter),
            revenue = sum(total_amount), total_hours = sum(total_hours))
ggplot(data = taxi_empty_month, aes(x = total_hours, y = revenue)) + geom_point()
fit <- lm(revenue ~ total_hours, data = taxi_empty_month)
summary(fit)
```

Interestingly, a higher proportion of working time off the meter doesn't necessarily mean less revenue.  This could be because longer trips (i.e. to/from the airport) could mean more empty time but are high-priced.

```{r}
ggplot(ml_best, aes(x = (total_empty_time/3600)/total_hours, y = total_revenue)) + geom_point() +
  geom_smooth()
```

Shall we try some machine learning, perhaps?

```{r, echo = FALSE}
strat_indiv_ml <- strat_indiv %>% 
  mutate(total_num_trips = total_trips) %>% 
  spread(payment_type, total_trips, fill = 0) %>%
  group_by(hack_license) %>% 
  summarize(mean_pickup_hour = mean(pickup_hour)
            , total_meter_hours = sum(total_time)/3600
            , total_distance = sum(total_distance)
            , total_num_passengers = sum(total_num_passengers)
            , total_trips = sum(total_num_trips)
            , CRD = sum(CRD)/total_trips
            , CSH = sum(CSH)/total_trips
            , total_revenue = sum(total_revenue)) %>% 
  mutate(rev_level = ifelse(total_revenue>9135, "high", 
                            ifelse(total_revenue<6276, "low", "mid"))) %>% 
  left_join(taxi_empty_month, by = "hack_license")
  
fit <- rpart(rev_level ~ .-total_revenue -hack_license -revenue, data = strat_indiv_ml, method = "class")
prp(fit)
```

This decision tree reveals three paths to high revenue:
1. Some distance (1218 miles < x < 1819 miles), many hours on the meter (>=132) ---> "meter method"
2. Lots of distance (>=1819 miles), many hours working (>=204) ---> "hour method"
3. Lots of distance (>=1819 miles), many trips (>=603) --> "trip method"

This begs the question, which path is the most lucrative?

```{r}
ml_best <- strat_indiv_ml %>% 
  filter(rev_level == "high") %>% 
  mutate(win_method = ifelse(total_distance >= 1218 & total_distance < 1819 & 
                               total_meter_hours >= 132, "t_meter", 
                        ifelse(total_distance >= 1819 & total_hours >=204, "t_hour",
                          ifelse(total_distance >= 1819 & total_hours < 204 & total_trips >= 603
                                 , "t_trip", "unknown")))) %>% 
  filter(win_method!="unknown") 

total_method <- ml_best %>% 
  group_by(win_method) %>% 
  summarize(count = n())
total_method

total_rev <- ml_best %>% 
  group_by(win_method) %>% 
  summarize(total_revenue = sum(total_revenue)) %>% 
  left_join(total_method, by = "win_method") %>% 
  mutate(rev_per_driver = total_revenue/count)
total_rev
```

So, in our sample of 222 high-performing taxi drivers, it looks like the hour method produced the most revenue per driver, closely followed by the meter method and the trip method. Let's compare the hours worked with each of these groups, in order to isolate the strategy with the least cost (assuming cost = hours working).

```{r, echo = FALSE}
ml_best %>% 
  select(hack_license, total_hours, win_method) %>% 
  ungroup() %>% 
  arrange(total_hours)

ggplot(ml_best, aes(x = total_hours, y = total_revenue, color = win_method, group = win_method)) + 
  geom_point() + xlab("Hours Worked") + ylab("Total Revenue") + ggtitle("High Performing NYC Cab Drivers, January 2013")
```

This graph is what we'd expect: the trip group spends fewer hours working, the meter group spends more hours working, and the hour group is much larger than the other two, accounting for the greater variance. It doesn't seem that there's one method that gives a particularly great return on investment (time).  

With long trips in mind, look at airport pickups: what percentage of cab drivers spend time picking up at the airport?

```{r}
airports <- taxi_strat %>% 
  mutate(airport_name = ifelse(pickup_longitude > -73.889 & pickup_longitude < -73.855
                               & pickup_latitude > 40.767 & pickup_latitude < 40.786, "LGA",
                        ifelse(pickup_longitude > -73.823 & pickup_longitude < -73.748
                               & pickup_latitude > 40.621 & pickup_latitude < 40.663, "JFK",
                               "none"))) 
airport_indiv <- airports %>% 
  group_by(hack_license, airport_name) %>% 
  summarize(count = n(), total_revenue = sum(total_amount)) %>% 
  spread(airport_name, count, fill = 0) %>% 
  group_by(hack_license) %>% 
  summarize(total_revenue = sum(total_revenue), JFK = sum(JFK), LGA = sum(LGA), none = sum(none)) %>% 
  mutate(airport_trips = JFK + LGA, total = JFK + LGA + none, airport_percent = airport_trips/total,
         JFK_percent = JFK/airport_trips)

ggplot(aes(x = airport_percent, y = total_revenue), data = airport_indiv) + geom_point() + xlim(0,.2) +
  geom_smooth(se = F)

airport_indiv %>% 
  filter(airport_percent == 0)

# About 10% of cabbies did not pick up anyone from an airport in January.

ggplot(data = airports_indiv, aes(x = JFK_percent, y = total_revenue)) + geom_point()

# what about airport pickups by date?

airport_date <- airports %>% 
  group_by(pickup_date, airport_name) %>% 
  summarize(count = n()) %>% 
  spread(airport_name, count) %>% 
  mutate(airport = JFK + LGA) 

ggplot(data = airport_date, aes(x = pickup_date, y = airport)) + geom_line() + 
  xlab("Pickup Date") + ylab ("Number of Airport Pickups") + ggtitle("Airport Pickups in January")
  
```

It doesn't look like more airport pickups necessarily means more revenue.  Similarly, there doesn't seem to be much of a difference in revenue between the two airports (although pickups from JFK are flat rate and pickups from LGA are not).

```{r}
airport_hours <- airports %>% 
  filter(airport_name!="none") %>% 
  group_by(pickup_hour) %>% 
  summarize(count = n(), total_revenue = sum(total_amount))
  
bp <- ggplot(data = airport_hours)
bp_count <- bp + geom_line(aes(x = pickup_hour, y = count)) + ggtitle("Airport Pickup Counts")
bp_rev <- bp + geom_line(aes(x = pickup_hour, y = total_revenue)) + ggtitle("Airport Pickup Revenue")
```

Looks like prime time for an airport pickup is around 4pm. Is this the case for pickups in general?

```{r}
times <- taxi_strat %>% 
  group_by(pickup_hour) %>% 
  summarize(count = n(), total_revenue = sum(total_amount))
  
bp2 <- ggplot(data = times)
bp2_count <- bp2 + geom_line(aes(x = pickup_hour, y = count)) + ggtitle("Pickup Counts")
bp2_rev <- bp2 + geom_line(aes(x = pickup_hour, y = total_revenue)) + ggtitle("Pickup Revenue")

grid.arrange(bp_count, bp2_count, ncol = 1)
grid.arrange(bp_rev, bp2_rev, ncol = 1)
```

Notice that general pickups seem to dip at around 4pm, while airport pickups rise.  So, the few cabbies that do pickups at 4pm go to the airport.  We see a similar shape with the revenue graphs.

```{r}
airport_percent <- airport_hours %>% 
  left_join(times, by = "pickup_hour") %>% 
  mutate(count_percent = count.x/count.y, rev_percent = total_revenue.x/total_revenue.y)

bp3 <- ggplot(data = airport_percent)
bp3_count <- bp3 + geom_line(aes(x = pickup_hour, y = count_percent)) + 
  ggtitle("Percentage of Airport Pickups")
bp3_rev <- bp3 + geom_line(aes(x = pickup_hour, y = rev_percent)) + 
  ggtitle("Percentage of Airport Pickup Revenue")
  
grid.arrange(bp3_count, bp3_rev)
```

These two graphs tell a similar story: at 5am and 4pm, we see an increase in the percentage of pickups at airports.  The 5am peak is understandable: early flights are arriving, and there are probably fewer potential passengers on the streets so early.  The 4pm peak is a little more surprising, however.  Not sure what to make of this as of yet.




